### 2025-03-26 sports-cognitive-psychology.qmd
20250326 - expert performance

Automaticity involves fast and effortless processing that requires little or no focused attention.
The more automatic a process is, the less likely a person is to be consciously aware of it.
For instance, breathing is a fairly automatic process—people can breathe even when they are not concentrating on breathing.
Automaticity helps free up resources for more demanding tasks.
Football players and others often face situations requiring multitasking—i.e., performing two (or more) tasks simultaneously.
For instance, a Quarterback must look for open receivers, monitor approaching defenders, move to open space when necessary, and make an accurate, well-timed throw.

@@

Almost all cell phone bans restrict the use of handheld cell phones while driving but do not restrict the use of hands-free devices.
This is implicitly based on the assumption that any impairment to driving while talking on a cell phone is due to the use of a handheld device.
However, an alternative possibility is that driving impairment with cell phone use is due to dividing one's attention.
If this is the case, it would impact driving with both handheld and hands-free devices.

A study by @Redelmeier1997 found that 25% of a sample of 699 accidents involved drives that had spoken on a cell phone within 10 minutes prior to the accident.
The use of handheld and hands-free phone were equally associated with accidents suggesting divided attention is the problem, not the use of a peripheral device.
However, this association was observed from a [correlational design](#sec-correlationalStudy); thus, we cannot conclude that cell phone use directly caused the accidents.
Thus, we need studies to examine the effects of multitasking using [experimental designs](#sec-experiment) to establish causality.

@Strayer2001 conducted an experimental study on multitasking.
For the first task, the tracking task, participants tracked the position of a moving target on a computer screen.
Every 10–20 seconds, the moving target flashed either red or green.
If the target turned red (such as one might see when a traffic light turns red), the participant was supposed to push a "brake button", which might simulate the act of needing to brake while driving.
If the target turned green (such as one might see when a traffic light turns green), the participant was not supposed to do anything.
During a dual-task phase, the participant was supposed to perform a second task while completing the tracking task.
For the second task, participants were randomly assigned to one of four conditions: 1) conversing with an experimenter using a hand-held mobile device, 2) conversing with an experimenter using a hands-free device, 3) listening to a radio program of their choice, or 4) listening to a book on tape.
The researchers found that the probability of missing a color change more than doubled for the cell phone group when performing both the tracking and conversation tasks simultaneously (relative to performing only the tracking task).
Reaction time also increased for the cell phone group when performing the dual tasks compared to when performing the single tracking task.
However, there was no difference—in terms of the probability of missing a color change or reaction time—between the dual tasks and single task for the radio or book-on-tape control groups, suggesting that listening, by itself, does not interfere with the tracking task—rather, it appears that actively having a conversation (that includes talking and listening) interferes with tracking changes.
Moreover, there was no difference between handheld or hands-free phones in terms of the probability of missing a color change or reaction time, suggesting that it was not the use of a hands-occupying device that caused their weaker performance in the tracking task—it was the act of dividing one's attention across tasks (i.e., conversation task and tracking task) that led to weaker performance in tracking changes.

Many people talk on the phone while driving.
What does it mean that people are (somehow) able to perform these two tasks simultaneously?
It suggests that those who use the cell phone while driving are treating one of the tasks—driving—as an automatic process and not giving it focused attention.
Evidence suggests that talking on the phone while driving increases the risk of accident to levels comparable to driving drunk [i.e., a block alcohol level above the legal limit; @Redelmeier1997].
The cognitive task of talking on the phone uses attentional resources that would otherwise be used for driving a car.

@Strayer2004a also replicated their findings using a driving simulator instead of a tracking task.
The authors found that use of a cell phone led to 18% slower brake onset time, 12% greater following distance, 17% longer half-recovery time (the time to recover 50% of the speed that was lost during braking), and twice as many rear-end collisions—all despite no general differences in speed.
Although older participants were poorer overall in driving performance, the effects of cell phone use did not differ for younger (~20 years of age) and older (~70 years of age) adults, suggesting that the driving performance for both younger and older adults is impaired by cell phone use.

The findings described above might suggest that more attention leads to better performance.
However, this is not always the case.
@Beilock2001 found that once a task becomes automatic, focusing too much on it can actually hinder performance.
When a task is highly practiced or automatic, excessive attention can hurt performance.
That is, more attention can lead to worse performance.
This is called "choking under pressure" and is based on explicit monitoring theory.
Highly practiced tasks might include tasks like putting in golf, pitching a baseball, throwing/kicking a football, etc.

By contrast, for tasks that require conscious control, decision-making, and real-time adjustments, more focused attention tends to lead to better performance.
Tasks that require real-time adjustment include tasks such as avoiding unexpected hazards while driving or trying to avoid rushers while operating as a Quarterback.
In sum, devoting explicit attention to a task appears to improve performance when the task requires deliberate cognitive processing but can hurt performance when the task is highly automatic.

>>>>

Automaticity involves fast and effortless processing that requires little or no focused attention.
The more automatic a process is, the less likely a person is to be consciously aware of it.
For instance, breathing is a fairly automatic process—people can breathe even when they are not concentrating on breathing.
Chewing food can also be automatic.
Automaticity helps free up resources for more demanding tasks.
Football players and others often face situations requiring multitasking—i.e., performing two (or more) tasks simultaneously.
For instance, a Quarterback must look for open receivers, monitor approaching defenders, move to open space when necessary, and make an accurate, well-timed throw.

@@

Almost all cell phone bans restrict the use of handheld cell phones while driving but do not restrict the use of hands-free devices.
This is implicitly based on the assumption, from the *peripheral-interference hypothesis*, that any impairment to driving while talking on a cell phone is due to the use of a handheld device.
However, an alternative possibility, the *attentional hypothesis*, posits that that driving impairment with cell phone use is due to dividing one's attention.
If this is the case, it would impact driving with both handheld and hands-free devices.

A study by @Redelmeier1997 found that 25% of a sample of 699 accidents involved drives that had spoken on a cell phone within 10 minutes prior to the accident.
The use of handheld and hands-free phone were equally associated with accidents suggesting divided attention is the problem, not the use of a peripheral device.
However, this association was observed from a [correlational design](#sec-correlationalStudy); thus, we cannot conclude that cell phone use directly caused the accidents.
It may be, for instance, that risk-taking individuals or individuals in a strongly emotional state may be more likely to both a) have car accidents and b) to use cell phones while driving, which might account for such an association.
Thus, we need studies to examine the effects of multitasking using [experimental designs](#sec-experiment) to establish causality.

@Strayer2001 conducted an experimental study on multitasking.
For the first task, the tracking task, participants tracked the position of a moving target on a computer screen.
Every 10–20 seconds, the moving target flashed either red or green to simulate a traffic signal.
If the target turned red (such as one might see when a traffic light turns red), the participant was supposed to push a "brake button", which might simulate the act of needing to brake while driving.
If the target turned green (such as one might see when a traffic light turns green), the participant was not supposed to do anything.
During a dual-task phase, the participant was supposed to perform a second task while completing the tracking task.
For the second task, participants were randomly assigned to one of four conditions: 1) conversing with an experimenter using a hand-held mobile device, 2) conversing with an experimenter using a hands-free device, 3) listening to a radio program of their choice, or 4) listening to a book on tape.
The researchers found that the probability of missing a color change (i.e., a simulated traffic signal) more than doubled for the cell phone group when performing both the tracking and conversation tasks simultaneously (relative to performing only the tracking task).
Reaction time also increased for the cell phone group when performing the dual tasks compared to when performing the single tracking task.
However, there was no difference—in terms of the probability of missing a color change or reaction time—between the dual tasks and single task for the radio or book-on-tape control groups, suggesting that listening, by itself, does not interfere with the tracking task—rather, it appears that actively having a conversation (that includes talking and listening) interferes with tracking changes.
Moreover, there was no difference between handheld or hands-free phones in terms of the probability of missing a color change or reaction time, suggesting that it was not the use of a hands-occupying device that caused their weaker performance in the tracking task—it was the act of dividing one's attention across tasks (i.e., conversation task and tracking task) that led to weaker performance in tracking changes.
Thus, the findings were consistent with the attentional hypothesis, and suggest that divided attention can interfere with performance—in addition to any deficits incurred due to peripheral factors such as handling the phone while dialing.

Many people talk on the phone while driving.
What does it mean that people are (somehow) able to perform these two tasks simultaneously?
It suggests that those who use the cell phone while driving are treating one of the tasks—driving—as an automatic process and not giving it focused attention.
Evidence suggests that talking on the phone while driving increases the risk of accident to levels comparable to driving drunk [i.e., a blood alcohol level above the legal limit; @Redelmeier1997].
The cognitive task of talking on the phone uses attentional resources that would otherwise be used for driving a car.

@Strayer2004a also replicated their findings using a driving simulator, which is a more [ecologically valid](#sec-ecologicalValidity) assessment of driving performance than the tracking task.
In the study, using the driving simulator, participants drove on a highway in both single-task (i.e., driving only) and dual-task (i.e., driving while talking on a cell phone) conditions.
Participants drove while following a car that braked at various times.
The authors found that use of a cell phone led drivers to show a 18% slower brake onset time, 12% greater following distance, 17% longer half-recovery time (the time to recover 50% of the speed that was lost during braking), and twice as many rear-end collisions—all despite no general differences in speed.
Although older drivers were poorer overall in driving performance, the effects of cell phone use did not differ for younger (~20 years of age) and older (~70 years of age) drivers, suggesting that the driving performance for both younger and older adults is impaired by cell phone use.
Indeed, the reaction time of younger drivers who were talking on a cell phone was equivalent to the reaction time of older drivers who were not using a cell phone.

The findings described above might suggest that more attention leads to better performance.
Based on that assumption, we should tell a Quarterback who has thrown multiple interceptions during a game to pay more attention to his throwing mechanics when throwing the ball, right?
However, it is not always the case that devoting more attention to a task leads to better performance.
There are two primary competing theories for "choking" in pressure situations in sports.
"Choking" refers to underperforming relative to one's ability [@Beilock2001].
Distraction theories posit that the pressure situation creates a distracting environment that shifts one's attention to cues that are irrelevant to the task, such as one's worries about the situation and its consequences [@Beilock2001].
An alternative possibility, based on explicit monitoring theory, is that pressure situations raise one's self-awareness and anxiety about performing well, which increases one's attention to the step-by-step procedures involved in skill processes, which may disrupt well-learned or proceduralized performance [@Beilock2001].
@Beilock2001 found that, among automatic tasks (e.g. putting in golf among highly trained individuals), focusing too much on the task can actually hinder performance, consistent with explicit monitoring theory.
When a task is highly practiced or automatic, excessive attention can hurt performance.
That is, more attention can lead to worse performance.
This is called "choking under pressure" and is based on explicit monitoring theory.

Explicit monitoring theories posit that pressure situations raises one's self-awareness and anxiety about performing well, which increases one's attention to the stop-by-step procedures involved in skill processes [@Beilock2001].
For highly practiced experts, automatic skill tasks might include tasks like hitting/putting a golf ball, pitching a baseball, throwing/kicking a football, shooting free throws in basketball, shooting arrows in archery, etc.
Sports players often refer to highly practiced tasks as involving "muscle memory."

By contrast, for tasks that require conscious control, decision-making, and real-time adjustments, more focused attention tends to lead to better performance.
Tasks that require real-time adjustment include tasks such as avoiding unexpected hazards while driving or trying to avoid rushers while operating as a Quarterback.
Unpredictable events occur when driving and when operating as a Quarterback; when unpredictable things happen and one's attentions is diverted, one's performance is likely to be impaired [@Strayer2001].
In sum, devoting explicit attention to a task appears to improve performance when the task requires deliberate cognitive processing but can hurt performance when the task is highly automatic.
